Global patterns of forest autotrophic carbon fluxes

### 2021-02-23 manuscript/manuscript draft.Rmd
final read-through
#129

Forest C fluxes decrease with latitude [e.g., @cramer_comparing_1999; @zhao_improvements_2005; @luyssaert_co_2007; @gillman_latitude_2015; @li_mapping_2019]. 
However, studies have differed in their conclusions regarding the shape of this relationship -- quite possibly because of lack of standardization with respect to methodology and stand history. 
C flux and allocation vary with stand age, disturbance, and management [@de_lucia_forest_2007; @yu_high_2014; @fernandez-martinez_spatial_2014; @simova_enigma_2017], making clear latitudinal patterns difficult to discern without standardization of the dataset.
Studies agree that gross primary productivity ($GPP$) increases continuously with decreasing latitude and is indisputably highest in tropical forests [@luyssaert_co_2007; @beer_terrestrial_2010; @jung_global_2011; @badgley_terrestrial_2019; @li_mapping_2019]. 
However, this relationship is more ambiguous for subsidiary fluxes. 
Some studies have suggested that net primary productivity ($NPP$), or its aboveground portion ($ANPP$), exhibits a less distinct increase from temperate to tropical forests [@luyssaert_co_2007] -- or even a decrease [@huston_global_2009, but see Gillman et al., 2015].
A shallower increase in $NPP$ than in $GPP$ with decreasing latitude would align with the suggestion that tropical forests tend to have low carbon use efficiency [$CUE$= $NPP$/$GPP$; @de_lucia_forest_2007; @malhi_productivity_2012; @anderson-teixeira_carbon_2016] but contrast with recent findings of the opposite pattern [@collalti_forest_2020].
Such differences among C fluxes in their relationship to latitude have profound implications for our understanding of the C cycle and its climate sensitivity [e.g., @collalti_forest_2020].
However, until recently the potential to compare latitudinal trends across C fluxes has been limited by lack of a large database with standardization for methodology, stand history, and management [@anderson-teixeira_forc_2018].

Latitudinal gradients in forest C flux rates, along with altitudinal gradients [@girardin_net_2010; @malhi_variation_2017; @mullerlandau_patterns_2020], are driven primarily by climate, which is a significant driver of C fluxes across broad spatial scales [@cramer_comparing_1999; @luyssaert_co_2007; @cleveland_relationships_2011; @wei_forest_2010; @mullerlandau_patterns_2020].
However, there is little consensus as to the shapes of these relationships or the best predictor variables.
The majority of studies have focused on exploring the relationships of C fluxes to mean annual temperature ($MAT$) and precipitation ($MAP$), which are the most commonly reported site-level climate variables. 
C fluxes increase strongly with $MAT$ on the global scale, but whether they saturate or potentially decrease at higher temperatures remains disputed. 
Some studies have detected no deceleration or decline in $GPP$ [@luyssaert_co_2007], $NPP$ [@schuur_productivity_2003], or root respiration [$R_{root}$; @piao_forest_2010; @wei_forest_2010] with increasing $MAT$.
In contrast, others have found evidence of saturation or decline of C flux in the warmest climates: @luyssaert_co_2007 found $NPP$ saturating at around 10$^\circ$C MAT; @larjavaara_temperature_2012 found that increases in $GPP$ saturate at approximately 25$^\circ$C $MAT$, and @sullivan_long-term_2020 found that, within the tropics, woody stem productivity ($ANPP_{stem}$) decreases at the highest maximum temperatures.
C fluxes generally saturate at high levels of $MAP$, though the saturation points identified vary widely [e.g., ~1000 - 2,445 mm yr^-1^; @wei_forest_2010; @schuur_productivity_2003]. 
Interactions between $MAT$ and $MAP$ may also influence productivity [@beer_terrestrial_2010; @yu_high_2014]; within the tropics, there is a positive interaction between $MAT$ and $MAP$ in shaping $ANPP$, such that temperature has a positive effect on productivity in moist climates, but a negative effect in dry climates [@taylor_temperature_2017]. 
There is also evidence that C fluxes also respond to climate variables such as seasonality of temperature and precipitation [@wagner_climate_2016], cloud cover [@taylor_temperature_2017], solar radiation [@beer_terrestrial_2010; @fyllas_solar_2017], and potential evapotranspiration [@kerkhoff_plant_2005]; however, these are not typically assessed in global-scale analyses of annual forest C flux.

@@

\renewcommand{\arraystretch}{2}
```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
hypothesis_table[1,1] <- paste0(hypothesis_table[1,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(2),"$^{,}$", footnote_marker_number(3)) #Luyssaert, Gillman et al. (2015)", "Simova and Storch (2017)"
hypothesis_table[2,1] <- paste0(hypothesis_table[2,1], footnote_marker_number(1),"$^{,}$", footnote_marker_number(4)) #Luyssaert, Huston & Wolverton
hypothesis_table[3,1] <- paste0(hypothesis_table[3,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(5),"$^{,}$",footnote_marker_number(6),"$^{,}$",footnote_marker_number(7)) ## Luyssaert, Schuur, Piao, Wei
hypothesis_table[4,1] <- paste0(hypothesis_table[4,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(4),"$^{,}$",footnote_marker_number(7)) ## Luyssaert, Schuur, Wei
hypothesis_table[5,1] <- paste0(hypothesis_table[5,1], footnote_marker_number(8),"$^{,}$",footnote_marker_number(9)) ### Taylor, muller Landau
hypothesis_table[7,1] <- paste0(hypothesis_table[7,1], footnote_marker_number(10)) ## smith et al. 2020
hypothesis_table[8,1] <- paste0(hypothesis_table[8,1], footnote_marker_number(11),"$^{,}$",footnote_marker_number(12)) ### fyllas nemani
hypothesis_table[10,1] <- paste0(hypothesis_table[10,1], footnote_marker_number(13),"$^{,}$",footnote_marker_number(14)) ## wagner 2016 + 2014
hypothesis_table[11,1] <- paste0(hypothesis_table[11,1], footnote_marker_number(15),"$^{,}$",footnote_marker_number(16),"$^{,}$", footnote_marker_number(17),"$^{,}$", footnote_marker_number(18)) ## and fernandez martinez
hypothesis_table[12,1] <- paste0(hypothesis_table[12,1], footnote_marker_number(16),"$^{,}$", footnote_marker_number(17))
hypothesis_table[13,1] <- paste0(hypothesis_table[13,1], footnote_marker_number(17))
hypothesis_table[15,1] <-paste0(hypothesis_table[15,1], footnote_marker_number(18)) ##fernandez martinez
kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Summary of review questions, corresponding expectations based on previous studies (when applicable), and results. Statistically signficant support for/ rejection of hypotheses is indicated by checkmarks/ X's, and '-' indicates no significant relationship. Parentheses indicate partial overall support or rejection of hypotheses across all fluxes considered. Flux variables are defined in Table 2.", escape = F) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12) %>%
  pack_rows("Q1. How do C fluxes vary with latitude?", 1, 2,  colnum =7, hline_before = FALSE) %>%
  pack_rows("Q2. How do C fluxes vary with mean annual temperature (MAT) and precipitation (MAP)?", 3, 5,  colnum = 7, hline_before = TRUE) %>%
  pack_rows("Q3. How are C fluxes related to other annual climate variables?", 6, 8,  colnum = 7, hline_before = TRUE) %>%
  pack_rows("Q4. How does seasonality influence annual C fluxes?", 9, 12,  colnum = 7, hline_before = TRUE) %>%
  pack_rows("Q5. When standardised by growing season length, how do annual C fluxes vary with climate?", 13, 16, colnum = 7, hline_before = TRUE) %>%
  add_header_above(c(" ",  " ", "Forest autotrophic carbon fluxes" = 9, " ")) %>%
  #column_spec(2:11, width = "1.8cm") %>%
  kableExtra::landscape() %>%
  kableExtra::footnote(number = c("Luyssaert et al. (2007)", "Gillman et al. (2015)", "Simova and Storch (2017)", "Huston & Wolverton (2009)", "Schuur (2003)", "Piao et al. (2010)", "Wei et al. (2010)", "Taylor et al. (2017)", "Muller-Landau et al. (2020)", "Smith et al. (2020)", "Fyllas et al. (2017)", "Nemani et al. (2003)", "Wagner et al. (2014)" , "Wagner et al. (2016)", "Malhi (2012)", "Michaletz et al. (2014)", "Chu et al. (2016)", "Fernandez-Martinez et al. (2014)"), threeparttable = TRUE, footnote_as_chunk = TRUE)
```

@@

We found no evidence of systematic variation in C allocation with latitude or climate (Fig. S3). Of 12 relationships tested (3 ratios among C flux variables regressed against latitude, $MAT$, $MAP$ and temperature seasonality), none were significant (all p>0.05).

@@

Forest C fluxes decline with temperature seasonality (Table 1, *Q4*; Fig. 4e), as is to be expected given that fluxes are minimal during winters. 
A temperature-defined growing season length correlated with global-scale variation in annual C flux [Table 1, *Q5*; Fig. 4f; see also @churkina_spatial_2005], consistent with the idea that the latitudinal gradient in carbon flux is attributable more to shorter growing seasons at high latitudes than to inherently lower rates of photosynthesis or respiration by high-latitude forests [@enquist_adaptive_2007; @fu_maximum_2019]. 
However, we find evidence that, within the growing season, climate still plays an important role in shaping C fluxes, as indicated by a number of positive correlations between monthly mean flux during the growing season and growing season temperature, solar radiation, and $PET$ (Table 1, Figs. S8-S9). This suggests that, while trees in high-latitude forests have adaptations to maximize photosynthesis at low temperatures [@helliker_subtropical_2008; @huang_air_2019], such adaptations are not sufficient to yield growing season fluxes comparable to those of tropical forests.
Thus, we reject the hypothesis that there is no relationship between C flux per month of the growing season and growing season climatic conditions [Table 1, *Q5*; @kerkhoff_plant_2005; @enquist_adaptive_2007; @michaletz_convergence_2014].
Rather, annual C flux is shaped by both growing season length and the climate of peak growing season months [@chu_does_2016; @fu_maximum_2019]. 
Given strong co-variation between growing season length and $MAT$ [Fig. S2; @chu_does_2016], accurately partitioning this variation will require data on intra-annual variation in C flux coupled with a higher-precision metric of growing season length than the monthly-resolution metric used here [e.g., based on leaf phenology or C exchange, *sensu* @fu_maximum_2019]. 
@fu_maximum_2019 find that global-scale geographic variation in annual $NEE$ is driven more strongly by growing season length than by carbon uptake rates within the growing season, whereas interannual variation in $NEE$ and $GPP$ at any given site appears to be driven predominantly by the maximum rate of C uptake, as opposed to growing season length [@zhou_explaining_2016; @fu_maximum_2019; @zani_increased_2020].
Further analysis of interannual variation in C fluxes in relation to climate will be valuable in disentangling how seasonality shapes broad geographic patterns in forest C flux.

>>>>

Forest C fluxes decrease with latitude [e.g., @cramer_comparing_1999; @zhao_improvements_2005; @luyssaert_co_2007; @gillman_latitude_2015; @li_mapping_2019; @anderson-teixeira_carbon_nodate]. 
However, studies have differed in their conclusions regarding the shape of this relationship -- quite possibly because of lack of standardization with respect to methodology and stand history. 
C flux and allocation vary with stand age, disturbance, and management [@de_lucia_forest_2007; @yu_high_2014; @fernandez-martinez_spatial_2014; @simova_enigma_2017], making clear latitudinal patterns difficult to discern without standardization of the dataset.
Studies agree that gross primary productivity ($GPP$) increases continuously with decreasing latitude and is indisputably highest in tropical forests [@luyssaert_co_2007; @beer_terrestrial_2010; @jung_global_2011; @badgley_terrestrial_2019; @li_mapping_2019]. 
However, this relationship is more ambiguous for subsidiary fluxes. 
Some studies have suggested that net primary productivity ($NPP$), or its aboveground portion ($ANPP$), exhibits a less distinct increase from temperate to tropical forests [@luyssaert_co_2007] -- or even a decrease [@huston_global_2009, but see Gillman et al., 2015].
A shallower increase in $NPP$ than in $GPP$ with decreasing latitude would align with the suggestion that tropical forests tend to have low carbon use efficiency [$CUE$= $NPP$/$GPP$; @de_lucia_forest_2007; @malhi_productivity_2012; @anderson-teixeira_carbon_2016] but contrast with recent findings of the opposite pattern [@collalti_forest_2020].
Such differences among C fluxes in their relationship to latitude have profound implications for our understanding of the C cycle and its climate sensitivity [e.g., @collalti_forest_2020].
However, until recently the potential to compare latitudinal trends across C fluxes has been limited by lack of a large database with standardization for methodology, stand history, and management [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_nodate].

Latitudinal gradients in forest C flux rates, along with altitudinal gradients [@girardin_net_2010; @malhi_variation_2017; @mullerlandau_patterns_2020], are driven primarily by climate, which is a significant driver of C fluxes across broad spatial scales [@cramer_comparing_1999; @luyssaert_co_2007; @cleveland_relationships_2011; @wei_forest_2010; @mullerlandau_patterns_2020].
However, there is little consensus as to the shapes of these relationships or the best predictor variables.
The majority of studies have focused on exploring the relationships of C fluxes to mean annual temperature ($MAT$) and precipitation ($MAP$), which are the most commonly reported site-level climate variables. 
C fluxes increase strongly with $MAT$ on the global scale, but whether they saturate or potentially decrease at higher temperatures remains disputed. 
Some studies have detected no deceleration or decline in $GPP$ [@luyssaert_co_2007], $NPP$ [@schuur_productivity_2003], or root respiration [$R_{root}$; @piao_forest_2010; @wei_forest_2010] with increasing $MAT$.
In contrast, others have found evidence of saturation or decline of C flux in the warmest climates: @luyssaert_co_2007 found $NPP$ saturating at around 10$^\circ$C MAT, @larjavaara_temperature_2012 found that increases in $GPP$ saturate at approximately 25$^\circ$C $MAT$, and @sullivan_long-term_2020 found that, within the tropics, woody stem productivity ($ANPP_{stem}$) decreases at the highest maximum temperatures.
C fluxes generally saturate at high levels of $MAP$, though the saturation points identified vary widely [e.g., ~1000 - 2,445 mm yr^-1^; @wei_forest_2010; @schuur_productivity_2003]. 
Interactions between $MAT$ and $MAP$ may also influence productivity [@beer_terrestrial_2010; @yu_high_2014]; within the tropics, there is a positive interaction between $MAT$ and $MAP$ in shaping $ANPP$, such that temperature has a positive effect on productivity in moist climates, but a negative effect in dry climates [@taylor_temperature_2017]. 
There is also evidence that C fluxes also respond to climate variables such as seasonality of temperature and precipitation [@wagner_climate_2016], cloud cover [@taylor_temperature_2017], solar radiation [@beer_terrestrial_2010; @fyllas_solar_2017], and potential evapotranspiration [@kerkhoff_plant_2005]; however, these are not typically assessed in global-scale analyses of annual forest C flux.

@@

\renewcommand{\arraystretch}{2}
```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
hypothesis_table[1,1] <- paste0(hypothesis_table[1,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(2),"$^{,}$", footnote_marker_number(3)) #Luyssaert, Gillman et al. (2015)", "Simova and Storch (2017)"
hypothesis_table[2,1] <- paste0(hypothesis_table[2,1], footnote_marker_number(1),"$^{,}$", footnote_marker_number(4)) #Luyssaert, Huston & Wolverton
hypothesis_table[3,1] <- paste0(hypothesis_table[3,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(5),"$^{,}$",footnote_marker_number(6),"$^{,}$",footnote_marker_number(7)) ## Luyssaert, Schuur, Piao, Wei
hypothesis_table[4,1] <- paste0(hypothesis_table[4,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(4),"$^{,}$",footnote_marker_number(7)) ## Luyssaert, Schuur, Wei
hypothesis_table[5,1] <- paste0(hypothesis_table[5,1], footnote_marker_number(8),"$^{,}$",footnote_marker_number(9)) ### Taylor, muller Landau
hypothesis_table[7,1] <- paste0(hypothesis_table[7,1], footnote_marker_number(10)) ## smith et al. 2020
hypothesis_table[8,1] <- paste0(hypothesis_table[8,1], footnote_marker_number(11),"$^{,}$",footnote_marker_number(12)) ### fyllas nemani
hypothesis_table[10,1] <- paste0(hypothesis_table[10,1], footnote_marker_number(13),"$^{,}$",footnote_marker_number(14)) ## wagner 2016 + 2014
hypothesis_table[11,1] <- paste0(hypothesis_table[11,1], footnote_marker_number(15),"$^{,}$",footnote_marker_number(16),"$^{,}$", footnote_marker_number(17),"$^{,}$", footnote_marker_number(18)) ## and fernandez martinez
hypothesis_table[12,1] <- paste0(hypothesis_table[12,1], footnote_marker_number(16),"$^{,}$", footnote_marker_number(17))
hypothesis_table[13,1] <- paste0(hypothesis_table[13,1], footnote_marker_number(17))
hypothesis_table[15,1] <-paste0(hypothesis_table[15,1], footnote_marker_number(18)) ##fernandez martinez
kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Summary of review questions, corresponding expectations based on previous studies (when applicable), and results. Statistically signficant support for/ rejection of hypotheses is indicated by checkmarks/ X's, whereas '-' indicates no significant relationship. Parentheses indicate partial overall support or rejection of hypotheses across all fluxes considered. Flux variables are defined in Table 2.", escape = F) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12) %>%
  pack_rows("Q1. How do C fluxes vary with latitude?", 1, 2,  colnum =7, hline_before = FALSE) %>%
  pack_rows("Q2. How do C fluxes vary with mean annual temperature (MAT) and precipitation (MAP)?", 3, 5,  colnum = 7, hline_before = TRUE) %>%
  pack_rows("Q3. How are C fluxes related to other annual climate variables?", 6, 8,  colnum = 7, hline_before = TRUE) %>%
  pack_rows("Q4. How does seasonality influence annual C fluxes?", 9, 12,  colnum = 7, hline_before = TRUE) %>%
  pack_rows("Q5. When standardised by growing season length, how do annual C fluxes vary with climate?", 13, 16, colnum = 7, hline_before = TRUE) %>%
  add_header_above(c(" ",  " ", "Forest autotrophic carbon fluxes" = 9, " ")) %>%
  #column_spec(2:11, width = "1.8cm") %>%
  kableExtra::landscape() %>%
  kableExtra::footnote(number = c("Luyssaert et al. (2007)", "Gillman et al. (2015)", "Simova and Storch (2017)", "Huston & Wolverton (2009)", "Schuur (2003)", "Piao et al. (2010)", "Wei et al. (2010)", "Taylor et al. (2017)", "Muller-Landau et al. (2020)", "Smith et al. (2020)", "Fyllas et al. (2017)", "Nemani et al. (2003)", "Wagner et al. (2014)" , "Wagner et al. (2016)", "Malhi (2012)", "Michaletz et al. (2014)", "Chu et al. (2016)", "Fernandez-Martinez et al. (2014)"), threeparttable = TRUE, footnote_as_chunk = TRUE)
```

@@

We found no evidence of systematic variation in C allocation with latitude or climate (Fig. S3). Of twelve relationships tested (three ratios among C flux variables regressed against latitude, $MAT$, $MAP$ and temperature seasonality), none were significant (all p>0.05).

@@

Forest C fluxes decline with temperature seasonality (Table 1, *Q4*; Fig. 4e), as is to be expected given that fluxes are minimal during winters. 
A temperature-defined growing season length correlated with global-scale variation in annual C flux [Table 1, *Q5*; Fig. 4f; see also @churkina_spatial_2005], consistent with the idea that the latitudinal gradient in carbon flux is attributable more to shorter growing seasons at high latitudes than to inherently lower rates of photosynthesis or respiration by high-latitude forests [@enquist_adaptive_2007; @fu_maximum_2019]. 
However, we find evidence that, within the growing season, climate still plays an important role in shaping C fluxes, as indicated by a number of positive correlations between monthly mean flux during the growing season and growing season temperature, solar radiation, and $PET$ (Table 1, Figs. S8-S9). This suggests that, while trees in high-latitude forests have adaptations to maximize photosynthesis at low temperatures [@helliker_subtropical_2008; @huang_air_2019], such adaptations are not sufficient to yield growing season fluxes comparable to those of tropical forests.
Thus, we reject the hypothesis that there is no relationship between C flux per month of the growing season and growing season climatic conditions [Table 1, *Q5*; @kerkhoff_plant_2005; @enquist_adaptive_2007; @michaletz_convergence_2014].
Rather, annual C flux is shaped by both growing season length and the climate of peak growing season months [@chu_does_2016; @fu_maximum_2019]. 
Given strong co-variation between growing season length and $MAT$ [Fig. S2; @chu_does_2016], accurately partitioning their influence will require data on intra-annual variation in C flux coupled with a higher-precision metric of growing season length than the monthly-resolution metric used here [e.g., based on leaf phenology or C exchange, *sensu* @fu_maximum_2019]. 
@fu_maximum_2019 find that global-scale geographic variation in annual $NEE$ is driven more strongly by growing season length than by carbon uptake rates within the growing season, whereas interannual variation in $NEE$ and $GPP$ at any given site appears to be driven predominantly by the maximum rate of C uptake, as opposed to growing season length [@zhou_explaining_2016; @fu_maximum_2019; @zani_increased_2020].
Further analysis of interannual variation in C fluxes in relation to climate will be valuable in disentangling how seasonality shapes broad geographic patterns in forest C flux.
